---
title: "IntegrativeCox Example"
author: "Aaron J. Molstad and Rohit K. Patra"
date: "Updated on January 28th, 2022"
output:
  pdf_document: default
  html_document: default
---
```{r, echo=FALSE}
functionspath <- "/Users/aaron/Documents/GitHub/IntegrativeCox/Functions/"
```

In this document, we provide a short example of how to (1) format data for integrative survival analysis with our method and (2) how to implement our method. First, let us generate $J = 12$ cancer datasets in the manner described in Section 5 of the manuscript. 
```{r setup, include=TRUE, cache=TRUE}
library(Matrix)
library(glmnet)
library(MASS)
library(survival)
set.seed(1)

J <- 12
Ns <- rep(c(1250, 1350, 1450), 4)
nreps <- 100
p <- 50
r <- 6
quan.cens <- .35
kappa <- seq(2000, 2110, by = 10)

genCoxDat <- function(uu, p, Ns, quan.cens){
  
  set.seed(1)
  SigmaX <- matrix(0, p, p)
  for(j in 1:p){
    for(k in 1:p){
      SigmaX[j,k] <- .7^(abs(j-k))
    }
  }
  X <- list(NA)
  for(j in 1:J){
    X[[j]] <- mvrnorm(n = Ns[j], mu = rep(0, p), 
                      SigmaX, tol = 1e-06, empirical = FALSE)
  }
  
  simulGomp <- function(datIndex, lambda, alpha, beta, rateC){
    
    N <- dim(X[[datIndex]])[1]
    x <- as.matrix(X[[datIndex]])
    v <- runif(n=N)
    Tlat <- (1/alpha)*log(1 - (alpha*log(v))/(lambda*exp(x%*%beta)))
    
    # censoring times
    if(datIndex%%3 == 0){  
      temp <- quantile(Tlat, quan.cens+.2)
    } else {
      temp <- quantile(Tlat, quan.cens)
    }  
    
    C <- rexp(n=N, rate=1/temp)
    
    # follow-up times and event indicators
    time <- pmin(Tlat, C)
    status <- 1*(Tlat <= C)
    
    # data set
    list("id"=1:N, "time"=time, "status"=status, "X"=x, "Tlat" = Tlat, "C" = C, "linPred" = x%*%beta)
  }
  
  
  get_beta <- function(sigma.temp){
    beta <- matrix(0, nrow=p, ncol=J)
    nonzeroes <- sample(1:p, 20, replace=FALSE)
    temp <- svd(matrix(rnorm(r*J), nrow=r, ncol=J))$v*(sqrt(2)/sqrt(r))
    beta[nonzeroes, ] <- matrix(runif(20*r, 1, 2)*
                                  sample(c(-1, 1), 20*r, replace=TRUE), nrow=20)%*%t(temp)
    return(beta)
  }
  
  sigma.temp <- NULL
  beta <- get_beta(sigma.temp)
  
  dat <- list(NA)
  for(kk in 1:J){
    alpha <- pi/(600*sqrt(6))
    lambda <- alpha*exp(- 0.5772 - alpha*kappa[kk])
    dat[[kk]] <- simulGomp(datIndex = kk, lambda=lambda, alpha=alpha, beta=beta[,kk], rateC=0.1)
  }
  
  return(list("dat" = dat, "beta" = beta, "SigmaX" = SigmaX))
  
}

simDat <- genCoxDat(uu = uu, p = p, Ns = Ns, quan.cens = quan.cens)
beta <- simDat$beta
dat <- simDat$dat
SigmaX <- simDat$SigmaX
simDat <- NULL


# ----------------------------------------------------
# Split into training, validation, and testing sets
# ----------------------------------------------------
datVal <- list(NA)
datTest <- list(NA)

for(j in 1:J){
  
  ValInds <- (100*((j-1)%%3 + 1) + 1):(100*((j-1)%%3 + 1) + 150)
  TestInds <- (100*((j-1)%%3 + 1) + 151):Ns[j]
  datVal[[j]] <- list(
    "X" = dat[[j]]$X[ValInds,],
    "time" = dat[[j]]$time[ValInds],
    "status" = dat[[j]]$status[ValInds],
    "Tlat" = dat[[j]]$Tlat[ValInds],
    "linPred" = dat[[j]]$linPred[ValInds])
  
  datTest[[j]] <- list(
    "X" = dat[[j]]$X[TestInds,],
    "time" = dat[[j]]$time[TestInds],
    "status" = dat[[j]]$status[TestInds],
    "Tlat" = dat[[j]]$Tlat[TestInds],
    "linPred" = dat[[j]]$linPred[TestInds])
  
  dat[[j]]$X <- dat[[j]]$X[-c(TestInds, ValInds),]
  dat[[j]]$time <- dat[[j]]$time[-c(TestInds, ValInds)]
  dat[[j]]$status <- dat[[j]]$status[-c(TestInds, ValInds)]
  
}

```
Now, we have three types of datasets: training, testing and validation ($\texttt{dat}, \texttt{datVal},$ and $ \texttt{datTest}$). Each is a list of length $J = 12$, and has subject id, event time (time), status (alive or deceased at event time), the predictor matrix $X$, and some other information which is not used for model fitting. 
```{r, include=TRUE, cache=TRUE}
length(dat)
str(dat[[1]])
```
Next, we load the functions and fit the model using $\texttt{IntCox}$ from the script $\texttt{LRCox.R}$. Note that one must first set the appropriate path $\texttt{functionspath}$. To shorten computing time, we will only consider $s \in \{5,10, 15\}$ and $r \in \{4, 6\}$. This will take a few minutes. To track progress, you may set $\texttt{quiet}$ or $\texttt{silent}$ equal to $\texttt{FALSE}$.
```{r, include=TRUE, cache=TRUE}
# ------- load functions
source(paste(functionspath, "LRCox.R", sep=""))
sourceCpp(paste(functionspath, "updateBeta.cpp", sep=""))

# ------- fit model for multiple (s,r) combinations
fit <- IntCox(svec = c(5, 10, 15), rvec = c(4,6), dat = dat, mu = 0.1, quiet = TRUE, silent = TRUE, rho0 = 100)
```
Examining the output, we see that $\texttt{fit}$ has three elements: $\texttt{beta, s, r}$. The array $\texttt{beta}$ is $p \times J \times s^\star \times r^\star$ where $s^\star$ is the number of candidate $s$ and $r^\star$ is the number of candidate $r$. Now, let us check the validation partial log-likelihood to determine which pair of tuning parameters is best. 

```{r, include=TRUE, cache=TRUE}
valerrsOurs <- array(0, dim=c(length(fit$s), length(fit$r),J))
for(kk in 1:(length(fit$s))){
  for(jj in 1:length(fit$r)){
    for(ll in 1:J){
      valerrsOurs[kk,jj,ll] <- coxnet.deviance(y=Surv(datVal[[ll]]$time, datVal[[ll]]$status), pred = datVal[[ll]]$X%*%fit$beta[,ll,kk,jj])
    }
  }
}

out <- which(apply(valerrsOurs, c(1,2),sum) == min(apply(valerrsOurs, c(1,2),sum)), arr.ind=TRUE)
betaLR <- fit$beta[,,out[1,1], out[1,2]]
cat("Tuning parameters selected by validation set:" , "\n")
cat("s = ", fit$s[out[1,1]], ", r = ", fit$r[out[1,2]], "\n")

```
We can examine the regression coefficient estimate: 
```{r, include = TRUE, cache=TRUE}
image(betaLR)
```

Similarly, let us extra the left singular vectors of the coefficient matrix estimate so that we may construct the factors for, say, the first population. 
```{r, include = TRUE, cache=TRUE}
U <- svd(betaLR)$u
u.fit <- U[,1:fit$r[out[1,2]]]
X1factors <- datTest[[1]]$X%*%u.fit
dim(X1factors)
```
Now, let us fit a model to the testing data using the estimated factors from the training data. 
```{r}
library(survival)
summary(coxph(Surv(datTest[[1]]$time, datTest[[1]]$status)~X1factors))
```
We can see that the factors lead to relatively high concordance, and all tests suggest they are significantly associated with survival. 